Assessing primitives performance\\ on multi-stage execution

### 2018-07-01 paper.tex
latest version before the bootstrap thingie

\title{Assessing primitives performance\\ on multi-stage \st{execution} implementations}
%\subtitle{Subtitle Text, if any\titlenote{with optional subtitle note}}

@@

On top of its virtual machine, the client language can use a set of \emph{primitives}\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation. Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method. We distinguish two kinds of primitives: \todo{motivate the distinction. See the algorithmic primitives paper. Is there a reference stating the different types of primitives ? }
\begin{itemize}
	\item \emph{Essential primitives} cannot be performed in any other way. A high-level object-oriented language without primitives can move values from one variable to another, but cannot add two integers together. Many arithmetic and comparison operations between numbers are primitives. Some primitives allow one to communicate with I/O devices such as the disk, the display and the keyboard.
	\item \emph{Optional primitives} exist only to make the system run faster \cite{blueBook}. They can be implemented in the client language directly, making its implementation as a primitive optional, but they are often implemented directly in the VM to improve performance.
\end{itemize}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve. \todo{a ref would be nice}
	\item There is a big performance gap between the baseline performance and the peak performance. \todo{a ref would be nice}
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
\end{itemize}

@@

\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the extended ASCII standard. \todo{we should not use this term. Maybre prefer Latin-1 standard} WideStrings encode each character in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (\st{ASCII or }Unicode) or using different orders, for example, the case insensitive order.

\paragraph{Primitive specification.} The String comparison optional primitive takes two or three parameters. The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters). If no order is specified, the VM assumes the byte ordering (\st{ASCII} \sk{Unicode} order) is the order to compare against, and compares directly the bytes of the ByteStrings. If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order. Having the order optional allows to have quicker string comparison in the common case where the \st{ASCII} \sk{Unicode} order is used. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

In the paper we will focus on the comparison of two strings with the default \st{ASCII}\sk{Unicode} ordering and measure the performance only of this case. The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non- \sk{Unicode }ASCII are considered uncommon.

@@

\paragraph{\legBase.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level\footnote{The bytecode compiler moves the upper limit of the loop ahead of the loop, so (length1 min: length2) is not compute at each iteration.}. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII \sk{Unicode} ordered comparison, we show the code only of baselineCompareWith:.

@@

%Vspce to fit all on page.
\vspace{-0.1cm}
	\paragraph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). 
\vspace{-0.1cm}
	\paragraph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
\vspace{-0.1cm}
	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation, \todo {} \st{which arguably can be used to evaluate the complexity level of each implementation.}
\vspace{-0.1cm}
	\paragraph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we count all the Slang implementations as well as the Cog's RTL one.
\vspace{-0.1cm}
	\paragraph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
\vspace{-0.1cm}
	\paragraph{Optimization potential:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
\vspace{-0.1cm}
	\paragraph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\vspace{0.2cm}

@@

\todo{mention baseline here}
The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

The loop overhead is measured and removed from the measurements (We made sure the JIT did not remove the dead loop to measure it). The two compared strings are equal as this situation corresponds to the worst scenario for most implementations.
\todo{explain how the relative speed up is computed} %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters. In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions. In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop. Figure~\ref{fig:bench} and Table \ref{tab:detail_bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.
The relative speed-up is obtained by dividing the baseline results by the execution times. Both the figure and the table shows bootstrapped standard errors. \todo{add ref for bootstrap ?}

@@

\section{Discussion and related work}
\todo{primitives in other system. (generalize more after Rpython. }

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitives is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain. \todo{emphasize the difference more}

@@

\acks
\todo{add acknowledgments}
Eliot, Levente
Acknowledgments, if needed.

>>>>

\title{Assessing primitives performance\\ on multi-stage execution}
%\subtitle{Subtitle Text, if any\titlenote{with optional subtitle note}}

@@

On top of its virtual machine, the client language can use a set of \emph{primitives}\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation. Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method. We distinguish two kinds of primitives: 
%\todo{motivate the distinction. See the algorithmic primitives paper. Is there a reference stating the different types of primitives ? }
\begin{itemize}
	\item \emph{Essential primitives} cannot be performed in any other way. A high-level object-oriented language without primitives can move values from one variable to another, but cannot add two integers together. Many arithmetic and comparison operations between numbers are primitives. Some primitives allow one to communicate with I/O devices such as the disk, the display and the keyboard.
	\item \emph{Optional primitives} exist only to make the system run faster \cite{blueBook}. They can be implemented in the client language directly, making its implementation as a primitive optional, but they are often implemented directly in the VM to improve performance.
\end{itemize}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve. %\todo{a ref would be nice}
	\item There is a big performance gap between the baseline performance and the peak performance. %\todo{a ref would be nice}
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
\end{itemize}

@@

\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the Latin-1 (ISO 8859-1) standard. WideStrings encode each character in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (Latin-1) or using different orders, for example, the case insensitive order.

\paragraph{Primitive specification.} The String comparison optional primitive takes two or three parameters. The two first operands are the two ByteStrings to compare and the third operand, optional, is the order table (a ByteArray of size 256 encoding the order of characters). If no order is specified, the VM assumes the byte ordering (Latin-1 order) is the order to compare against, and compares directly the bytes of the ByteStrings. If an order is specified, for example a case insensitive order (the entries for the 41 and 61 in the order byte array, respectively characters A and a in the ASCII table, both answer the same value), the primitive compares the bytes through the indirection order. Having the order optional allows to have quicker string comparison in the common case where the Latin-1 order is used. The primitive answers a negative integer, zero, or a positive integer as the first parameter is less than, equal to, or greater than the second parameter.

In the paper we will focus on the comparison of two strings with the default Latin-1 ordering and measure the performance only of this case. The goal of this work was to improve the performance of common string operations, typically string equality which internally uses the primitive, and the cases where the order is non-Latin-1 are considered uncommon.

@@

\paragraph{\legBase.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level\footnote{The bytecode compiler moves the upper limit of the loop ahead of the loop, so (length1 min: length2) is not compute at each iteration.}. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on Latin-1 ordered comparison, we show the code only of baselineCompareWith:.

@@

%Vspce to fit all on page.
\vspace{-0.1cm}
	\paragraph{Recompilation required:} Changing the implementation requires only Smalltalk bytecode recompilation (No), requires to recompile a plugin to a dynamic library if compiled as external or the recompilation of the full VM if compiled as internal (Plugin) or requires to recompile the full VM (Yes). 
\vspace{-0.1cm}
	\paragraph{Language familiar to developers:} Smalltalk developers are considered to be familiar with Smalltalk only, while Cog VM developers are considered being familiar with Smalltalk, Slang, C and Assembly code. Only Cog's JIT implementors are familiar with its RTL, which is a subset of the VM developers.  We use All if all communities are familiar with the language, VM if all VM developers are and VM-JIT if only a subset of VM developers are.
\vspace{-0.1cm}
	\paragraph{Number of lines of code:} We measured the number of lines of code of each implementation.
\vspace{-0.1cm}
	\paragraph{Number of implementations to maintain:} In each case, the client language has to implement a fall-back code in Smalltalk to be executed if the primitive fails. This code is maintained by the client language implementors, and not by the VM implementors. Hence we do not count the fall-back code in the number of implementation to maintain, nor the pure Smalltalk implementation. However, as a VM implementor, we count all the Slang implementations as well as the Cog's RTL one.
\vspace{-0.1cm}
	\paragraph{Stack switch overhead:} Yes if activating the primitive from the native code generated by the JIT requires to switch between the C and the client stack, inducing execution time overhead.
\vspace{-0.1cm}
	\paragraph{Optimization potential:}
	C cross-file inlining is not performed by default for internal plugins. External plugins don't have access to the whole VM API and have to go through an interface to do so, leading to an overhead, limiting the performance unless the string size is really high.
\vspace{-0.1cm}
	\paragraph{Control over the generated code:} The implementations written in Slang are compiled by the C compilers into native code. The generated code can therefore differ according the compilers and the optimizations they performed. On the contrary, the JIT generates the same native code whichever C compiler is used.
\vspace{0.2cm}

@@

%\todo{mention baseline here}
The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

The loop overhead is measured and removed from the measurements (We made sure the JIT did not remove the dead loop to measure it). The two compared strings are equal as this situation corresponds to the worst scenario for most implementations.
%\todo{explain how the relative speed up is computed} %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters. In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions. In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop. Figure~\ref{fig:bench} and Table \ref{tab:detail_bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.
%The relative speed-up is obtained by dividing the baseline results by the execution times. Both the figure and the table shows bootstrapped standard errors. \todo{add ref for bootstrap ?}

@@

\section{Discussion and related work}
%\todo{primitives in other system. (generalize more after Rpython. }

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitives is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain. %\todo{emphasize the difference more}

@@

\acks
We thank Levente Uzonyi for providing the benchmark template used in this article and Eliot Miranda for his advice on the implementation.
